Two Problems for Sophistication

### 2015-01-21 Facticity/Facticity/paper.tex
...

Adriaans' \emph{Facticity}, introduced in 2012\cite{adriaans2012facticity}, relaxes the model class to include all partial recursive functions. Adriaans also identifies the nickname problem discussed in Section~\ref{section:indices}.

Other authors have pointed out problems with sophistication. Antunes et al. \cite{antunes2009sophistication} note that the constant cutoff used by Koppel is a source of instability, a result in the spirit of this paper. They propose a different definition called \emph{coarse sophistication}. This version behaves very differently from regular sophistication, and is closer to computational depth. Vereshchagin \cite{vereshchagin2013algorithmic} also points out that the Kolmogorov minimal sufficient statistic is not always well-defined and suggests requiring that the finite set model should described using a \emph{total} function rather than any computable function. This yields a more robust sophistication that is nonetheless still susceptible to the issues discussed in this paper.
\section{Preliminaries}
Let $\B = \{0,1\}^*$. We deal with partial computable functions $\phi: \B \to \B$, which we will also call \emph{models}. A function is called \emph{prefix} if its domain, $\dom{f} = \{y : f(y) \neq \infty\}$ is a prefix free set, i.e. no string in $\dom{f}$ is a prefix of another. A function $f$ is \emph{total} if $\dom{f} = \B$.

@@

For the first argument, we take a generative perspective. We can generate data from a model in $\phi_i, i \in \K$ by feeding it random bits until it produces an output. We will call the resulting probability distribution $p_i(x) = \sum_{y:\phi_i(y) = x} 2^{-|y|}$. In this scenario, a sophistication could be called \emph{consistent} if, for data of sufficiently high complexity, it accurately reflects the complexity of that distribution from which it originated. Suppose that the data issampled from  universal distribution: let $\phi_u(\bar\imath y)=\phi_i(y)$ and sample from $p_u$ as described above. Then the initial random bits will determine the prefix encoded index $\bar\imath$ of the function $\phi_i$ that $\phi_u$ will subsequently emulate, and the remaining bits are used as inputs to $\phi_i$. We now ask, what should be the sophistication of the resulting data?

@@

The second reason we believe sophistication will elude definition is more technical. Given an input $x$ and some model class, say $\C$, consider the set of all possible two-part representations of $x$. Now, when the numbering of partial computable functions is changed, the codelength of the model part of all these representations will change. This is illustrated in the right graph in Figure~\ref{fig:diagram}. However, the Invariance Theorem expresses that this change is limited by a constant term that does not depend on $x$. So for sufficiently large $x$, the horizontal shift of all the representations will be relatively small. However, even this small shift can push some representations out of the acceptable region (indicated by the dashed line), and pull others in. This may lead to a different representation determining the sophistication, one whose \emph{total} codelength is close to what it was before, but whose \emph{model} codelength can be anywhere between $0$ and $C^\C(x)$. If such jumps can occur, the sophistication is not invariant. And while we cannot \emph{prove} in general that such jumps can always occur, there seems to be no reason to believe that they don't.

For this reason, we take a skeptical view of sophistication. Note that part of the theory is fine: there is nothing wrong with evaluating models for the data by comparing their two-part code lengths. In fact, the randomness deficiency $\delta(\phi_i,x)=-\log p_i(x)-C^\K(x\mid i)$ of a model $\phi_i$ has a direct statistical interpretation as a measure of counterevidence against the hypothesis that it generated the data---if $p_i$ truly generated the data, then the probability that the randomness deficiency is more than $k$ bits is less than $2^{-k}$ \cite{bloem2014safe}[Lemma~6]. In the Monet example above, this will allow us to disqualify the model expressing that the data is actually, say, a recording of Jazz music.

@@

\section{Proofs and further discussion}

@@

\begin{lemma}[Invariance of function complexity]
Let $\phi$ and $\psi$ be any two acceptable numberings. There exists a constant $c$ such that $\left| K_\phi(f) - K_\psi(f)\right | \leq c$ for all $f$. \label{lemma:invariance}
\end{lemma}
\begin{proof}
Let $g(i)$ be the function such that $\psi_i=\phi_{g(i)}$.
\begin{align*}
K_\phi(f) &= \min\left\{ K_\phi(i) : \phi_i= f\right\} 
\geq \min\left\{ K_\psi(i) : \phi_i= f\right\} - c\\
&= \min\left\{ K_\psi(i) : \psi_{g(i)}= f\right\} - c
\geq \min\left\{ K_\psi(i) : \psi_i= f\right\} - c' = K_\psi(f).
\end{align*}
The opposite inequality can be achieved by reversing $\phi$ and $\psi$. 
\end{proof}
This lemma shows that the complexity of a function does not depend on the choice of
enumeration by more than a constant term.

\begin{lemma}
  There is a faithful acceptable numbering.\label{lemma:faithful-numberings}
\end{lemma}
\begin{proof}
Let $\tn{div} \in \N$ be an index such that $\phi_{\tn{div}}(y)=\infty$ for all $y$. Define
  \[\psi_q=\begin{cases}
    \phi_{\phi_i(p)}&\tn{if $q$ can be written as $\bar\imath p$ and $\phi_i(p)<\infty$,}\\
    \phi_{i_\tn{div}}&\tn{otherwise.}\end{cases}
  \]
  To show that $\psi$ is faithful, pick any function $f$. Then
\[\begin{split}
C(f)&=\min\{C(i):\phi_i=f\} =\min\{\min\{|\bar a b|:\phi_a(b)=i\}:\phi_i=f\} \\
& =\min\{|\bar a b|:\phi_{\phi_a(b)}=f\}
 =\min\{|\bar a b|:\psi_{\bar a b}=f\}.
\end{split}\]
This shows there is a sufficiently small $\psi$ index.

To show that $\psi$ is acceptable, let $\phi_j$ denote the identity
function. Then a $\phi$-index $i$ can be mapped to a $\psi$-index
using the computable function $r(i)=\bar\jmath i$, so that
$\psi_{r(i)}(y)=\psi_{\bar\jmath i}(y)=\phi_i(y)$. For the reverse,
define $\phi_v(\bar\imath p, y)=\phi_{\phi_i(p)}(y)$. For fixed
$\bar\imath p$, the 
$s^n_m$-theorem \cite{kleene193notation} states that we can compute the $h$
such that $\phi_h(y)=\phi_v(\bar\imath p,y)$. Let $h(\bar\imath p)$
denote this index as a function of the program; further define
$h(q)=\tn{div}$ if $q$ cannot be expressed as $\bar\imath p$. By
construction $h$ is total and computable. To check that the mapping
returns the correct function, rewrite $\phi_{h(\bar\imath
  p)}(y)=\phi_v(\bar\imath p,y)=\phi_{\phi_i(p)}(y)=\psi_{\bar\imath p}(y)$.
\end{proof}

@@

\ineffprefix*
\begin{proof}
Let $\bar\imath y$ be any two-part representation for the data $x$, i.e. $\phi_i(y)=x$. Then construct an alternative two-part representation $\bar vi^* y$, where $i^*$ is the shortest $\psi_u$-program for $i$ such that $\phi_v(i^* y)=\phi_{\psi_u(i^*)}(y) = \phi_i(y)=x$. We compare the lengths of these two representations. Note that $|i^*|=K^\psi(i)$. Therefore,
\[
|\bar\imath p|-|\bar v i^* y| = |\bar\imath|-|\bar v| - |i^*| = |\bar\imath|-|\bar v|-K(i).
\]
By assumption there must be an $i_0$ such that the above expression is positive for all $i>i_0$. From this $i_0$ onwards, the second representation (using $v$) will have a shorter codelength, so $\bar\imath p$ cannot achieve the minimum in the definition of the sophistication. Consequently, the sophistication is bounded by $\s(x)<|\overline{\imath_0}|$ for all $x$. 
\end{proof}

@@

\begin{proof}
We define the numbering $\phi$ as follows:
\[\begin{cases}
\phi_0(p) = 1^r 0 D(p) \\
\phi_{1^r0i}(p) = \psi_i(p) \\
\phi_j(\cdot) = \infty &\text{if $j$ contains no zeroes.}
\end{cases}\]
We will show that under the $\phi$-enumeration, the best representation for $x$ using a model $f\in\M'$ is always better than the best representation using some $f\not\in\M'$.

First suppose $f\in\M'$. Then
\begin{align*}
K^\phi(f) &=\min\{|\bar\jmath q|:\phi_{\phi_j(q)}=f\}
\leq\min\{|\bar0 q|:\phi_{\phi_0(q)}=f\} \\ 
&=\min\{|\bar0 q|:\phi_{1^r0 D(q)}=f\}  
=|\bar 0|+\min\{|q|:\psi_{D(q)}=f\}
\leq K^\psi(f)+c+|\bar 0|,
\end{align*}
where the last inequality uses the first assumption.

Now assume that the best model $f$ for $x$ is not in $\M'$. 
Let $i$ be the index of $f$ with the shortest description, i.e. it achieves the minimum in $K^\phi(f)=\min\{K^\phi(i):\phi_i=f\}$.
There are two possibilities. Either $i=0$, in which case we have $K^\phi(f)\ge r$ because $\phi_0$ cannot output zero and all other $\phi$-programs are at least $r$ bits long.

Otherwise, we can bound
\begin{align*}
K^\phi(f)&=K^\phi(i)=K^\phi(1^r0j)\\
&\ge K^\phi(j)-c' \ge K^\psi(j)+r+1-c'\\
&=K^\psi(f)+r+1-c'.
\end{align*}
Now choose $r=c''+\max\{K^\psi(\phi_0),c'-1\}$. Then substitution yields, for both cases, 
$K^\phi(f)\ge c''+K^\psi(f)$.

>>>>

Antunes et al. \cite{antunes2009sophistication} note that the constant cutoff used by Koppel is a source of instability, a result in the spirit of this paper. They propose a different definition called \emph{coarse sophistication}. As we shall see, this version behaves very differently from regular sophistication, and is closer to computational depth. Vereshchagin \cite{vereshchagin2013algorithmic} suggests additional restrictions on models that are allowed to determine the sophistication, in hopes of curbing the instability issue. Adriaans' \emph{Facticity}, introduced in 2012\cite{adriaans2012facticity}, relaxes the model class to include all partial recursive functions. Adriaans also identifies the nickname problem discussed in Section~\ref{section:indices}.

\section{Preliminaries}
Let $\B = \{0,1\}^*$. We deal with partial computable functions $\phi: \B \to \B$, which we will also call \emph{models}. A function is called \emph{prefix} if its domain, $\dom{f} = \{y : f(y) \neq \infty\}$ is a prefix free set, i.e. no string in $\dom{f}$ is a prefix of another. A function $f$ is \emph{total} if $\dom{f} = \B$.

@@

For the first argument, we take a generative perspective. We can generate data from a model in $i \in \K$ by feeding it random bits until it produces an argument, we will call the resulting probability distribution for model $p_i(x) = \sum_{y:\phi_i(y) = x} 2^{-|y|}$. In this scenario, a sophistication could be called \emph{consistent} if, for data of sufficiently high complexity, it accurately reflects the complexity of that distribution from which it originated.\footnotemark Suppose that the data issampled from  universal distribution: let $\phi_u(\bar\imath y)=\phi_i(y)$ and sample from $p_u$ as described above. Then the initial random bits will determine the prefix encoded index $\bar\imath$ of the function $\phi_i$ that $\phi_u$ will subsequently emulate, and the remaining bits are used as inputs to $\phi_i$. We now ask, what should be the sophistication of the resulting data?

\footnotetext{While in general we do not wish to make any assumptions about the origin of the data, in some cases, it can make sense to do so as a sanity check.}

@@

The second reason we believe sophistication will elude definition is more technical. Given an input $x$ and some model class, say $\C$, consider the set of all possible two-part representations of $x$. Figure~\ref{fig:diagram} provides an illustration. Now, when the numbering of partial computable functions is changed, the codelength of the model part of all these representations will change. This is illustrated in the right graph in the figure. However, the Invariance Theorem expresses that this change is limited by a constant term that does not depend on $x$. So for sufficiently large $x$, the horizontal shift of all the representations will be relatively small. \emph{However}, even this small shift can push some representations out of the acceptable region (indicated by the dashed line), and pull others in. This may lead to a different representation determining the sophistication, one whose \emph{total} codelength is close to what it was before, but whose \emph{model} codelength can be anywhere between $0$ and $C^\C(x)$. If such jumps can occur, the sophistication is not invariant. And while we cannot \emph{prove} in general that such jumps can always occur, there seems to be no reason to believe that they don't.

For this reason, we take a skeptical view of sophistication. Note that \emph{part} of the theory is fine: there is nothing wrong with evaluating models for the data by comparing their two-part code lengths. In fact, the randomness deficiency $\delta(\phi_i,x)=-\log p_i(x)-C^\K(x\mid i)$ of a model $\phi_i$ has a direct statistical interpretation as a measure of counterevidence against the hypothesis that it generated the data---if $\phi_i$ truly generated the data, then the probability that the randomness deficiency is more than $k$ bits is less than $2^{-k}$ \cite{bloem2014safe}[Lemma~6]. In the Monet example above, this will allow us to disqualify the model expressing that the data is actually, say, a recording of Jazz music.

@@

\section{Proofs}

@@

\begin{lemma}[Invariance of function complexity]
Let $\psi$ and $\eta$ be any two acceptable numberings. There exists a constant $c$ such that $\left| C^{\K,\psi}(f) - C^{\K, \eta}(f)\right | \leq c$ for all $f \in \C$ and $\left| C^{\C,\psi}(f) - C^{\C, \eta}(f)\right | \leq c$ for all $f \in \C$. \label{lemma:invariance}
\end{lemma}
\begin{proof}
Let $g(i)$ be the function such that $\psi_i=\eta_{g(i)}$.
\begin{align*}
C^{\C,\psi}(f) &= \min\left\{ C^{\C,\psi}(i) : \psi_i= f\right\} 
\geq \min\left\{ C^{\C, \eta}(i) : \psi_i= f\right\} - c\\
&= \min\left\{ C^{\C, \eta}(i) : \eta_{g(i)}= f\right\} - c
\geq \min\left\{ C^{\C, \eta}(i) : \eta_i= f\right\} - c' \\
&= C^{\C, \eta}(f).
\end{align*}
The opposite inequality can be achieved by reversing $\psi$ and $\eta$. The same proof holds with model class $\K$.
\end{proof}
This lemma shows that the complexity of a function does not depend on the choice of
enumeration by more than a constant term.

\begin{lemma}
  There is a faithful acceptable numbering.\label{lemma:faithful-numberings}
\end{lemma}
\begin{proof}
Let $d \in \N$ be an index such that $\phi_d(y)=\infty$ for all $y$. Define
  \[\psi_q=\begin{cases}
    \phi_{\phi_i(p)}&\tn{if $q$ can be written as $\bar\imath p$ and $\phi_i(p)<\infty$,}\\
    \phi_d&\tn{otherwise.}\end{cases}
  \]
  To show that $\psi$ is faithful, pick any function $f$. Then
\[\begin{split}
C^{\C, \phi}(f)&=\min\{C^{\C, \phi}(i):\phi_i=f\} =\min\{\min\{|\bar a b|:\phi_a(b)=i\}:\phi_i=f\} \\
& =\min\{|\bar a b|:\phi_{\phi_a(b)}=f\}
 =\min\{|\bar a b|:\psi_{\bar a b}=f\}.
\end{split}\]
This shows there is a sufficiently small $\psi$ index.

To show that $\psi$ is acceptable, let $\phi_j$ denote the identity
function. Then a $\phi$-index $i$ can be mapped to a $\psi$-index
using the computable function $r(i)=\bar\jmath i$, so that
$\psi_{r(i)}(y)=\psi_{\bar\jmath i}(y)=\phi_i(y)$. For the reverse,
define $\phi_v(\bar\imath p, y)=\phi_{\phi_i(p)}(y)$. For fixed
$\bar\imath p$, the 
$s^n_m$-theorem \cite{kleene193notation} states that we can compute the $h$
such that $\phi_h(y)=\phi_v(\bar\imath p,y)$. Let $h(\bar\imath p)$
denote this index as a function of the program; further define
$h(q)=d$ if $q$ cannot be expressed as $\bar\imath p$. By
construction $h$ is total and computable. To check that the mapping
returns the correct function, rewrite $\phi_{h(\bar\imath
  p)}(y)=\phi_v(\bar\imath p,y)=\phi_{\phi_i(p)}(y)=\psi_{\bar\imath p}(y)$.
\end{proof}

@@

\ineffprefix*
\begin{proof}
Let $\bar\imath y$ be any two-part representation for the data $x$, i.e. $\phi_i(y)=x$. Then construct an alternative two-part representation $\bar vi^* y$, where $i^*$ is the shortest $\psi_u$-program for $i$ such that $\phi_v(i^* y)=\phi_{\psi_u(i^*)}(y) = \phi_i(y)=x$. We compare the lengths of these two representations. Note that $|i^*|=C^{\K,\phi}(i)$. Therefore,
\[
|\bar\imath y|-|\bar v i^* y| = |\bar\imath|-|\bar v| - |i^*| = |\bar\imath|-|\bar v|-C^{\K}(i).
\]
By assumption there must be an $i_0$ such that the above expression is positive for all $i>i_0$. From this $i_0$ onwards, the second representation (using $v$) will have a shorter codelength, so $\bar\imath p$ cannot achieve the minimum in the definition of the sophistication. Consequently, the sophistication is bounded by $s(x)<|\overline{\imath_0}|$ for all $x$. 
\end{proof}

@@

\begin{proof}
We define the numbering $\psi$ as follows:
\[\begin{cases}
\psi_0(p) = 1^r 0 D(p) \\
\psi_{1^r0i}(p) = \phi_i(p) \\
\psi_j(\cdot) = \infty &\text{if $j$ contains no zeroes.}
\end{cases}\]
We will show that under the $\psi$-enumeration, the best representation for $x$ using a model $m\in\M'$ is always better than the best representation using some $m\notin\M'$.

First suppose $m\in\M'$. Then
\begin{align*}
C^{\K,\psi}(m) &=\min\{|\bar\jmath q|:\psi_{\psi_j(q)}=m\}
\leq\min\{|\bar0 q|:\psi_{\psi_0(q)}=m\} \\ 
&=\min\{|\bar0 q|:\psi_{1^r0 D(q)}=m\}  
=\min\{|q|:\phi_{D(q)}=m\} + |\bar 0|\\
&\leq C^{\K,\phi}(m)+c+|\br{0}|,
\end{align*}
where the last inequality uses assumption (1).

Now assume that the best model $m$ for $x$ is not in $\M'$. 
Let $i$ be the index of $m$ with the shortest description, i.e. it achieves the minimum in $C^{\K, \psi}(m)=\min\left\{C^{\K,\psi}(i):\psi_i=m\right\}$.
There are two possibilities. Either $i = 0$, in which case we have $C^{\K,\psi}(m)\ge r$ because $\phi_0$ cannot output zero and all other $\psi$-programs are at least $r$ bits long.

Otherwise, we can bound
\begin{align*}
C^{\K,\psi}(m)&=C^{\K\psi}(i)=K^\phi(1^r0j)\\
&\ge K^\phi(j)-c' \ge K^\psi(j)+r+1-c'\\
&=K^\psi(f)+r+1-c'.
\end{align*}
Now choose $r=c''+\max\{K^\psi(\phi_0),c'-1\}$. Then substitution yields, for both cases, 
$K^\phi(f)\ge c''+K^\psi(f)$.
